Inference of a universal social scale and segregation measures using social connectivity kernels

### 2020-07-09 manuscript/main.tex
Incorporate changes except for measure-statistic terminology.

\title{``Inferring social kernels and segregation measures from ego-network data'' or ``Inferring social kernels and segregation measures with universal scale from ego-network data'' or ``Social kernels and segregation measures with universal scale: an application to ego-network data'' or ``Segregation measures with universal scale: an application to ego-network data''}
\author{Till Hoffmann}
\author{Nick S. Jones}
\affil{Department of Mathematics, Imperial College London}
\date{}

\begin{document}

@@

\begin{abstract} % max 200 words
    How people connect with one another is a fundamental question in the social sciences, and the resulting social networks can have a profound impact on our daily lives. Blau offered a powerful explanation: people connect with one another based on their positions in a social space. Yet a principled measure of social distance in society remains elusive.

    We use the connectivity kernel of conditionally-independent edge models to develop a family of segregation measures with desirable properties: they have an intuitive and universal scale (facilitating comparison across datasets and societies), are applicable to multivariate and mixed node attributes, and capture segregation at the level of individuals, pairs of individuals, and society as a whole. We show that the measures induce a (semi-)metric on Blau space (a space spanned by the attributes of the members of society) and provide maps of two societies.

    Under a Bayesian paradigm, we infer the parameters of the connectivity kernel from eleven ego network datasets collected in four surveys in the United Kingdom and United States. The importance of different dimensions of Blau space is similar across time and location, suggesting a macroscopically stable social fabric. Physical separation and age differences have the most significant impact on segregation within friendship networks with implications for intergenerational mixing and isolation in later stages of life.
\end{abstract}

@@

@@

Building on the successes of conditionally-independent edge models~\cite{Snijders2011} and, in particular, latent space models for social networks~\cite{Hoff2002,Handcock2007,Raftery2012}, we consider a generative model for social networks whose members occupy a multidimensional Blau space in \cref{sec:model}. We discuss desirable properties for social segregation measures, and, using the generative network model, we develop a suite of measures applicable to arbitrary attributes in \cref{sec:model-based-segregation}. The measures capture segregation at different scales: single individuals, pairs of individuals, and society as a whole. Because of their probabilistic foundations, the segregation measures have a universal scale, i.e.\ one unit of segregation has the same implications across different societies or at different times. We show that the measure for pairs of individuals is a semi-metric and can be used to measure distance in Blau space. We illustrate the measures with a simple example, and we show that it reduces to a well-known segregation measure if the attributes are univariate and categorical.

@@

@@

\subsection{Distance in Blau space\label{sec:segregation-distance}}

@@

In other words, the social separation is a true measure of \emph{distance} in the social space with a probabilistic interpretation if features are themselves measures of distance, including all the features we consider subsequently.
% TAH: may have to move that to the application section It remains a true metric for the standardised features we considered in \cref{sec:survey-application} because standardisation is an affine transform.
This observation puts Peter Blau's~\cite{Blau1977} hypothesis that ``the macrostructure of societies can be defined as a multidimensional space of social positions among which people are distributed and which affect their social relations'' on a sound statistical footing: \emph{fitting conditionally-independent edge models allows us to learn the metric of Blau space}. The metric has a universal scale, i.e.\ one unit of segregation has the same probabilistic meaning independent of the society under consideration, facilitating comparison across disparate datasets.

@@

\begin{figure}
    \includegraphics{embedding}
    \titlecaption{A lower-dimensional embedding of the inter-node distances reveals an interpretable social space.}{Panels~(a) and~(b) show the mean and standard deviation of ages as a function of the first embedding dimension as a solid line and a shaded region, respectively. Panels~(c) and~(d) show a scatter plot of respondents in a two-dimensional embedding space whose coordinates were obtained from the social separation $\varphi$ using multidimensional scaling. The colour of a marker indicates the respondent's ethnicity. The heat map represents a smoothed estimate of the social isolation $\phi$. The ``bands'' of individuals in panel~(c) correspond to different occupational statuses, such as employed or retired.\label{fig:segregation-embedding}}
\end{figure}

@@

We considered a generative model for social networks embedded in Blau space, a space spanned by the demographic attributes of members of society. We developed a family of segregation measures with a universal scale, facilitating comparison between datasets collected at different times or in different cultural contexts. Furthermore, the segregation measures are applicable to mixed attribute types, have a natural reference point, and an intuitive interpretation: the probability to form connections decreases with increasing segregation. For certain logistic connectivity kernels, the social separation is a metric for Blau space and allows us to quantify social distance in a principled fashion.

>>>>

\title{Inference of a universal social scale and segregation measures using social connectivity kernels}
\author{Till Hoffmann}
\author{Nick S. Jones}
\affil{Department of Mathematics, Imperial College London}
\date{}
%Buzz words -- inference, universal social scale, segregation measures, social kernel
\begin{document}

@@

\begin{abstract} % max 200 words
    How people connect with one another is a fundamental question in the social sciences, and the resulting social networks can have a profound impact on our daily lives. Blau offered a powerful explanation: people connect with one another based on their positions in a social space. Yet a principled measure of social distance, allowing comparison within and between societies, remains elusive.

    We use the connectivity kernel of conditionally-independent edge models to develop a family of segregation measures with desirable properties: they offer an intuitive and universal characteristic scale on social space (facilitating comparison across datasets and societies), are applicable to multivariate and mixed node attributes, and capture segregation at the level of individuals, pairs of individuals, and society as a whole. We show that the segregation measures induce a (semi-)metric on Blau space (a space spanned by the attributes of the members of society) and provide maps of two societies.

    Under a Bayesian paradigm, we infer the parameters of the connectivity kernel from eleven ego-network datasets collected in four surveys in the United Kingdom and United States. The importance of different dimensions of Blau space is similar across time and location, suggesting a macroscopically stable social fabric. Physical separation and age differences have the most significant impact on segregation within friendship networks with implications for intergenerational mixing and isolation in later stages of life.
\end{abstract}

@@

Peter Blau proposed that individuals connect with one another based on their positions in a high-dimensional space~\cite{Blau1977}, e.g.\ a space spanned by demographic attributes. With an accrual of large-scale survey data, we now have access to demographic and relational information internationally, for whole societies, and over time~\cite{McPherson2001,McPherson2006,Mossong2008}. Despite this wealth of data, we lack two important quantities: (a)~a natural notion of distance in this space, allowing us to determine how far apart individuals are in society, and (b)~a universal characteristic scale, allowing the distance between pairs of individuals in one society to be compared to another society with different social dimensions. We will argue that the probability of forming a friendship, intimately related to work on homophily by Blau and his successors~\cite{McPherson2001}, offers both a universal characteristicscale and a notion of distance.

@@

Building on the successes of conditionally-independent edge models~\cite{Snijders2011} and, in particular, latent space models for social networks~\cite{Hoff2002,Handcock2007,Raftery2012}, we consider a generative model for social networks whose members occupy a multidimensional Blau space in \cref{sec:model}. We discuss desirable properties for social segregation measures, and, using the generative network model, we develop a suite of measures applicable to arbitrary attributes in \cref{sec:model-based-segregation}. The measures capture segregation at different scales: single individuals, pairs of individuals, and society as a whole. Because of both their probabilistic foundations and their construction from the universal notion of a social tie, the segregation measures have a universal scale, i.e.\ one unit of segregation has the same implications across different societies or at different times. We show that the segregation measure for pairs of individuals is a semi-metric and can be used to quantify distance in Blau space. We illustrate the measures with a simple example, and we show that it reduces to a well-known segregation measure if the attributes are univariate and categorical.

@@

We have so far emphasised the desirability of metrics on social spaces with a universal characteristic scale (in the sense of being comparable between societies). After first developing social segregation measures, including a notion of social separation, we will formulate both a metric and a notion of scale in \cref{sec:segregation-distance}.

@@

\subsection{Distance and scale in Blau space\label{sec:segregation-distance}}

@@

In other words, the social separation is a true measure of \emph{distance} in the social space with a probabilistic interpretation if features are themselves measures of distance, including all the features we consider subsequently.
% TAH: may have to move that to the application section It remains a true metric for the standardised features we considered in \cref{sec:survey-application} because standardisation is an affine transform.
This observation puts Peter Blau's~\cite{Blau1977} hypothesis that ``the macrostructure of societies can be defined as a multidimensional space of social positions among which people are distributed and which affect their social relations'' on a sound statistical footing: \emph{fitting conditionally-independent edge models allows us to learn the metric of Blau space}. The metric has a universal scale: \emph{one unit of social separation has the same probabilistic meaning independent of the society under consideration}, facilitating comparison across disparate datasets. Even if two societies have different Blau space dimensions, e.g.\ a society might exist that strongly discriminates based on characteristics which are not found in other societies, the social separation between a pair of individuals has a common meaning.

@@

\begin{figure}
    \includegraphics{embedding}
    \titlecaption{A lower-dimensional embedding of the inter-node distances reveals an interpretable social space in the UK and US.}{Panels~(a) and~(b) show the mean and standard deviation of ages as a function of the first embedding dimension as a solid line and a shaded region, respectively. Panels~(c) and~(d) show a scatter plot of respondents in a two-dimensional embedding space whose coordinates were obtained from the social separation $\varphi$ using multidimensional scaling. The colour of a marker indicates the respondent's ethnicity. The heat map represents a smoothed estimate of the social isolation $\phi$. The ``bands'' of individuals in panel~(c) correspond to different occupational statuses, such as employed or retired.\label{fig:segregation-embedding}}
\end{figure}

@@

We considered a generative model for social networks embedded in Blau space, a space spanned by the demographic attributes of members of society. We developed a family of segregation measures with a universal scale (since they are based on the common notion of the probability of a social tie) facilitating comparison between datasets collected at different times or in different cultural contexts. Furthermore, the segregation measures are applicable to mixed attribute types, have a natural reference point, and an intuitive interpretation: the probability to form connections decreases with increasing segregation. For certain logistic connectivity kernels, the social separation is a metric for Blau space and allows us to quantify social distance in a principled fashion.
